Amortization of Sub-programs in Simulators

### 2019-09-08 paper/learning/notes.tex
updated nores and changed structure - cpp load is not working still...

\section{Sub-Programs within Simulators}

\begin{itemize}
  \item Talk about how within simulators we have nested-structures which are \emph{nested-programs}
  \item Define the notations used to define the processes within the simulator, in terms of the PPL framework
  \item Talk about the problems that such nested-programs cause to inference procedures
  \item Define the density of the simulator in terms of these sub-programs. 
\end{itemize}

@@

Rejection samplers are heavily used when designing simulators [ add SHERPA etc], due to their ease of use and ability to model 
any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella}

There are two different forms of rejection samplers found with in simulators and program code. 
In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 
sample from, but can evaluate.
We then define a proposal using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from.
Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
In the standard statistical setting, see Algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
Here $M$ is a constant that has a finite bound, $M \in (1, \infty)$ and the likelihood ratio $\frac{f(x)}{g(x)}$ is defined over the 
support of $X$. 
$M$ must also satisfy the following inequality $f(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $f(x)>0$. 
We construct this acceptance probability by sampling from a uniform distribution defined on the interval 
$(0,1)$, which the proof ultimately relies on to ensure that a sample generated from the rejection sampler directly 
corresponds to a sample from the true, normalized, target density $f(x)$. 
But, provided we can calculate $P_{\text{accept}}(\mathcal{D} \leq \mathcal{C})$ for the given density, or mass, function
 $\mathcal{D}$ and condition $\mathcal{C}$, we can also satisfy the standard proof with a little more work.
 However, as we will demonstrate, rejection samplers 
 need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
 More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
\begin{wrapfigure}{L}{0.40\textwidth}
 \begin{minipage}{0.40\textwidth}
  \null 
   \begin{algorithm}[H]
   \caption{ABC: Rejection Sampling}
   \label{alg:rsstats}
      \begin{algorithmic}
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
      \end{algorithmic}
    \end{algorithm}
  \end{minipage}%
\end{wrapfigure}

>>>>

\section{Sub-Programs within Simulators}
\label{sec:nested}
\begin{itemize}
  \item Talk about how within simulators we have nested-structures which are \emph{nested-programs}
  \item Define the notations used to define the processes within the simulator, in terms of the PPL framework
  \item Talk about the problems that such nested-programs cause to inference procedures
  \item Define the density of the simulator in terms of these sub-programs. 
\end{itemize}

@@

Rejection samplers allow from sampling from densities $p(X=x)$ by constructing an upper bound to $p(X=x)$
specified by some predicate expression that determines whether we accept or reject samples from a density proportional to the bounding envelope. 
The functions we choose to define the envelope are much simpler than $p(X=x)$ and easy to sample from. 
When designing rejection samplers we aim to reduce the rate of rejections and increase the rate of acceptance, for computational purposes.
However, designing efficient and expressive rejection samplers is challenging, leading to complex and inefficient constructions of 
rejection samplers in simulators, see appendix~\ref{app:exampleprograms} for examples.
Due to these inefficiencies, and additional complications that arise in nested structures, as described in Section~\ref{sec:nested}
we propose a scheme to amortize such processes and learn accurate surragotes for the individual sub-programs, which in many cases are complex rejection
sampling schemes.

% are heavily used when designing simulators,
%  due to their ease of use and ability to model 
% any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} 
\subsection{Types of rejection samplers}

There are essentially two different forms of rejection samplers found with in simulators and program code. 
In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 
sample from, but can evaluate.
We then define a proposal using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from.
Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
In the standard statistical setting, see Algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
Here $M$ is a constant that has a finite bound, $M \in (1, \infty)$ and the likelihood ratio $\frac{f(x)}{g(x)}$ is defined over the 
support of $X$. 
$M$ must also satisfy the following inequality $f(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $f(x)>0$. 
We construct this acceptance probability by sampling from a uniform distribution defined on the interval 
$(0,1)$, which the proof ultimately relies on to ensure that a sample generated from the rejection sampler directly 
corresponds to a sample from the true, normalized, target density $f(x)$. 
But, provided we can calculate $P_{\text{accept}}(\mathcal{D} \leq \mathcal{C})$ for the given density, or mass, function
 $\mathcal{D}$ and condition $\mathcal{C}$, we can also satisfy the standard proof with a little more work.
 However, as we will demonstrate, rejection samplers 
 need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
 More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
\begin{wrapfigure}{L}{0.40\textwidth}
 \begin{minipage}{0.40\textwidth}
  \null 
   \begin{algorithm}[H]
   \caption{ABC: Rejection Sampling}
   \label{alg:rsstats}
      \begin{algorithmic}
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
      \end{algorithmic}
    \end{algorithm}
  \end{minipage}%
\end{wrapfigure}
